---
title: " The University of Massachusetts Cashflows on Tuition Revenue"
author:
- affiliation: University Of Massachuessetts, Amherst
  name: Alain Duplan, Ariel Lutati
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    latex_engine: pdflatex
    template: tex_template.tex
  word_document: default
  html_document:
    df_print: paged
fontsize: 10pt
geometry: margin=.75in
fontfamily: mathpazo
abstract: \par The University of Massachusetts is a public institution providing collegiate education to tens of thousands of students spread out between its campuses statewide. Billions of dollars of revenue is accumulated by the institution each and every year. The use of numerous sources of income has resulted in a net gain of profit yearly. One of the most prominent sources is the net total of tuition and fees. Given as the main functionality of the UMass schools is provided higher education to pursuing students, the revenue made off tuition could be stated as the most important source, even if it may not be the biggest. However, one may question if other cash flows would affect this source. A good practice in business is to keep all income independent of one another to avoid a chain reaction of effects. However shifts in other sources of income could lead to the institution making decisions that affect the student body leading to an adjustment in revenue as more students may want to attend or leave or causing a revaluation of tuition costs. **We hypothesize that the revenue coming from tuition and fees for the University of Massachusetts is dependent on the revenue of its other cash flows.** Through the use of random foresting and linear regression we aimed to reject our hypothesis. Upon testing we weren’t able to formulate a model that was able to accurately predict revenue of tuition in fees, therefore correlation between the cash flows could not be drawn.

---

```{r, echo=FALSE, message=F, warning=F}
library(knitr)
opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readxl)
library(randomForest)
library(rlist)
library(Metrics)
library(TTR)
library(gridExtra)
library(xtable)
```


## **Introduction**

Amherst is the flagship campus of the University of Massachusetts. Being so it is one of the most populated colleges in the state. With more and more students wanting to pursue degrees, UMass, like most schools, is admitting more and more students each year. Initially this doesn't seem to be an issue as an extended student body would lead to greater student engagement and more income directly to the school to fund projects such as renovations. However, when the once cozy z-rooms are now becoming “economy triples,”  and the common rooms on every floor are now turning into quads and triples, we begin to question whether not the school is prepared to hold so many students.
\par When asked, a majority of people would say that UMass makes most of its income from tuition fees as it is a common misconception that the univerisity has direct control on the price of tuition. However, this is not necessarily true as tuition is set by the Department of Education and as UMass is a public institution, most students are sensitive with how much they pay for their tuition and fees. With national college debt on the rise, the growing cost of education worries the next generation on how they will afford it. Many pointed towards UMass’s habit of over enrolling students to be the cause of the increasing cost of tuition.
\par Taking into consideration that many students are now struggling to find rooms for next semester, and that overall student debt is at all time high, it is hard not to question whether there is a correlation between enrollment and tuition and fees. While controlling as many variables as possible we aim to find if enrollment is a determinant of the cost of attending UMass. 

## **Data**

\par The data we used was found on the University of Massachusetts office of the President website. Using the single audit reports of each year, parsed the cash flows for the years 2015 to 2019 into a single balance sheet. We focus on using only the net values of each category of  the cash flows, along with the overall total cash flows, and the changes compared the previous years.
\newline

```{r}
CashFlows <- read_excel("UMass Balance sheets.xlsx", sheet = 1)
#CashFlows <- as.data.frame(t(CashFlows))
original <- CashFlows
tuition <- CashFlows[c(2,18),]
#names(CashFlows)
CashFlows <- subset(CashFlows, Nets == "1")
CashFlows <- rbind(CashFlows, tuition)
kable(head(CashFlows[1:6], 12))
```

```{r}
train <- CashFlows[,2:4]
test <- CashFlows[,c(5:6)]
train <- as.data.frame(t(train))
test <- as.data.frame(t(test))

train.ts <- ts((as.numeric(as.vector(train$V12))), frequency = 1, start = c(2018, 1))
test.ts <- ts((as.numeric(as.vector(test$V12))), frequency = 1, start = c(2018, 1))

```
```{r}
seed.val <- 1234567
set.seed(seed.val)
```

\par We will also look deeper at cash flows from non financing activities as it was recently discovered that due to the recent coronavirus outbreak, UMass is using the revenue from grants to refund their students. So we would like to see if there was any correlation between the two.
\newline

```{r}
x <- original[c(2,3,16:20),]
x.train <- x[,2:4]
x.train <- as.data.frame(t(x.train))
kable(head(x[1:6], 7))
```

\par Upon initial inspection we see that the revenue of tution looks to be as if it is steadily decreasing over the years. The loss between 2015 and 2019 was almost 150 million dollars in fact.
\newline
```{r}
tuition <- as.data.frame(t(CashFlows))
 tuition.ts <- ts((as.numeric(as.vector(tuition[2:6,12]))), frequency = 1, start = c(2015, 1))
ts.plot(tuition.ts, xlab = "year", ylab ="amount in 1000 of dollars", main = "Tuition Over the Years")
```

## **Modeling**
\par To begin we will reserve the last 2 years of the dataframe as our testing set, and everything else as our training set.

\par We will first approach using  models linear regression models. When attempting to create a model using all the variables made available to us, the model would result is statistically insignificant coefficient estimates. So instead we approached this by making single variate regression models on all the variables we used in our data frame. As a result we found that 5 of the variables were significant on their own: sumplimental disclosures of non cash activities, net decrease in cash and noncash equivalents, cash flows from capital and other financing activities, grants and contracts, student orginzation agency transactions. 
\newline
```{r}
fit.3 <- lm(V12 ~V3, data = train)
kable(xtable(fit.3))
```
```{r}
fit.5 <- lm(V12 ~V5, data = train)
kable(xtable(fit.5))
```
```{r}
fit.11 <- lm(V12 ~V11, data = train)
kable(xtable(fit.11))
```
```{r}
x.2 <- lm(V1~V2,data=x.train)
kable(xtable(x.2))
```
```{r}
x.7 <- lm(V1~V7,data=x.train)
kable(xtable(x.7))
```

## **Results**


## **Conclusion**

\par We were able to draw a clear correlation between the rate of which enrollment increases and the rate of which costs increases. While controlling for not only inflation, but the effect of time, we were still able to predict the increase of cost to attend UMass within a roughly 5% interval. Though this is not a direct causation between the two, we can conclude that expected enrollment is at least considered when deciding the costs for next semester. The models weren't able to accurately account for business and economic cycles which can affect the price of which tuition is set. We are also only looking at purely the populous of UMass, not other factors such as staffing, programs, and resources provided by the school. With all the new buildings and renovations, there is a direct affect on room and board cost as it increases the value of life off campus. We must also keep in mind that UMass doesn’t directly choose their tuition as that is the decision of the Department of Education. 
\par Taking all of these into consideration, it remains undeniable that the relationship between the two holds true. There is in fact a correlation present; therefore, we are unable to reject our hypothesis. There is a significant relationship between the amount of students enrolled at the University of Massachusetts, Amherst and the cost of attending while staying on campus.
\newline



\newpage


## **References**


[“Reports &amp; Audits.” Reports &amp; Audits | University of Massachusetts Office of the President](www.umassp.edu/controller/reports)
\newline
\newline
[Thys, Fred. Study: Out-Of-State Students With Lower Test Scores Admitted To UMass Amherst More. WBUR, 29 May 2018](www.wbur.org/edify/2018/05/29/out-of-state-admissions-to-umass)
\newline
\newline
[“U.S. Bureau of Labor Statistics.” U.S. Bureau of Labor Statistics, U.S. Bureau of Labor Statistics, 28 Apr. 2020](www.bls.gov/)
\newline
\newline


