---
title: " The University of Massachusetts Cashflows on Tuition Revenue"
author:
- affiliation: University Of Massachuessetts, Amherst
  name: Alain Duplan, Ariel Lutati
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: pdflatex
    template: tex_template.tex
  word_document: default
fontsize: 10pt
geometry: margin=.75in
fontfamily: mathpazo
abstract: \par The University of Massachusetts is a public institution providing collegiate education to tens of thousands of students spread out between its campuses statewide. Billions of dollars of revenue is accumulated by the institution each and every year. The use of numerous sources of income has resulted in a net gain of profit yearly. One of the most prominent sources is the net total of tuition and fees. Given as the main functionality of the UMass schools is provided higher education to pursuing students, the revenue made off tuition could be stated as the most important source, even if it may not be the biggest. However, one may question if other cash flows would affect this source. A good practice in business is to keep all income independent of one another to avoid a chain reaction of effects. However shifts in other sources of income could lead to the institution making decisions that affect the student body leading to an adjustment in revenue as more students may want to attend or leave or causing a revaluation of tuition costs. **We hypothesize that the revenue coming from tuition and fees for the University of Massachusetts is dependent on the revenue of its other cash flows.** Through the use of random foresting and linear regression we aimed to reject our hypothesis. Upon testing we weren’t able to formulate a model that was able to accurately predict revenue of tuition in fees, therefore correlation between the cash flows could not be drawn.
---

```{r, echo=FALSE, message=F, warning=F}
library(knitr)
opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readxl)
library(randomForest)
library(rlist)
library(Metrics)
library(TTR)
library(gridExtra)
library(xtable)
library(tinytex)
```


## **Introduction**

The UMass system allows for a quality education for an economical price (for instate students). UMass's tuition and fees pricing is primarily decided by the Department of Education. Nonetheless, we wanted to explore for any internal factors within UMass' cashflows that could influence their revenue produced by tuition and fees. We believe that that if UMass acquires higher revenue in cashflows such as grants and contracts, operating, and non-operating activities, then quality of life on campus would increase consequently leading to higher revenue in tuition and fees. Likewise, if losses in cashflows increases, campus quality of life will decrease and thus, revenue made via tuition and fees will decrease.
\par Taking into consideration that overall student debt is at an all time high, it is hard not to question whether there is a correlation between other cashflows and tuition and fees. Given that students pay entirely towards tuition and fees. It would be within the students' best interests to understand what factors affect the students' largest contributer to academic finances and loan debt. 

## **Data**

\par The data we used was found on the University of Massachusetts office of the President website. Using the single audit reports of each year, parsed the cash flows for the years 2015 to 2019 into a single balance sheet. We focus on using only the net values of each category of  the cash flows, along with the overall total cash flows, and the changes compared the previous years.
\newline

```{r}
CashFlows <- read_excel("UMass Balance sheets.xlsx", sheet = 1)
#CashFlows <- as.data.frame(t(CashFlows))
original <- CashFlows
tuition <- CashFlows[c(2,18),]
#names(CashFlows)
CashFlows <- subset(CashFlows, Nets == "1")
CashFlows <- rbind(CashFlows, tuition)
kable(head(CashFlows[1:6], 12))
```

```{r}
train <- CashFlows[,2:4]
test <- CashFlows[,c(5:6)]
train <- as.data.frame(t(train))
test <- as.data.frame(t(test))

train.ts <- ts((as.numeric(as.vector(train$V12))), frequency = 1, start = c(2018, 1))
test.ts <- ts((as.numeric(as.vector(test$V12))), frequency = 1, start = c(2018, 1))

```
```{r}
seed.val <- 1234567
set.seed(seed.val)
```

\par We will also look deeper at cash flows from non financing activities as it was recently discovered that due to the recent coronavirus outbreak, UMass is using the revenue from grants to refund their students. So we would like to see if there was any correlation between the two.
\newline

```{r}
x <- original[c(2,3,16:20),]
x.train <- x[,2:4]
x.train <- as.data.frame(t(x.train))
kable(head(x[1:6], 7))
```

\par Upon initial inspection we see that the revenue of tution looks to be as if it is steadily decreasing over the years. The loss between 2015 and 2019 was almost 150 million dollars in fact.
\newline
```{r}
tuition <- as.data.frame(t(CashFlows))
 tuition.ts <- ts((as.numeric(as.vector(tuition[2:6,12]))), frequency = 1, start = c(2015, 1))
ts.plot(tuition.ts, xlab = "year", ylab ="amount in 1000 of dollars", main = "Tuition Over the Years")
```

## **Modeling**
<<<<<<< HEAD
\par To begin we will reserve the last 2 years of the dataframe as our testing set, and everything else as our training set.

\par Our first approach used linear regression models. When we created a model using all the variables made available to us, the model resulted in statistically insignificant coefficient estimates (meaning all null hypotheses were accepted). So instead we approached this by making single variate regression models on all the variables we used in our data frame. As a result we found that 5 of the variables were significant on their own: supplimental disclosures of non cash activities, net decrease in cash and noncash equivalents, cash flows from capital and other financing activities, grants and contracts, and  student organization agency transactions. 
\newline
```{r}
fit.3 <- lm(V12 ~V3, data = train)
kable(xtable(fit.3))
```
```{r}
fit.5 <- lm(V12 ~V5, data = train)
kable(xtable(fit.5))
```
```{r}
fit.11 <- lm(V12 ~V11, data = train)
kable(xtable(fit.11))
```
```{r}
x.2 <- lm(V1~V2,data=x.train)
kable(xtable(x.2))
```
```{r}
x.7 <- lm(V1~V7,data=x.train)
kable(xtable(x.7))
```

## **Results**


## **Conclusion**

\par We were unable to draw a clear correlation between tuition & fees and other cashflows. none of our models using significant factors produced an RMSE low enough to be accepted. Simply put, at first glance, factors such as Supplemental Disclosure Of Noncash Activities seemed to affect tuition and fees by about a \$933 USD decrease in tuition and fees per \$1000 increase. However, when the model's fit was eventually tested, it failed horribly by producing astronomically high RMSE values. This seems logical since we must also keep in mind that UMass doesn’t directly choose their tuition as that is the decision of the Department of Education. 
\par Taking all of these into consideration, it's undeniable that other cashflows preside no affect over UMass' revenue from tuition and fees. Therefore, we must reject our hypothesis. There is a  no significant relationship between other cashflows for the University of Massachusetts system and the cost of tuition and fees.
\newline



\newpage
=======

## **Results**
>>>>>>> 6b0da0d7457368234dc0023f3f3d2bb2cafab83e


## **References**


“Reports &amp; Audits.” Reports &amp; Audits | University of Massachusetts Office of the President
\newline
\newline
Thys, Fred. Study: Out-Of-State Students With Lower Test Scores Admitted To UMass Amherst More. WBUR, 29 May 2018
\newline
\newline
“U.S. Bureau of Labor Statistics.” U.S. Bureau of Labor Statistics, U.S. Bureau of Labor Statistics, 28 Apr. 2020
\newline
\newline


