---
title: "CheapMass"
output: html_document
author: "Alain Duplan & Ariel Lutati"
---

## set up

```{r, echo=FALSE, message=F, warning=F}
library(knitr)
opts_chunk$set(echo = FALSE)
library(tidyverse)
library(readxl)
library(randomForest)
library(rlist)
library(Metrics)
library(TTR)
library(gridExtra)
library(xtable)
```
```{r}
CashFlows <- read_excel("UMass Balance sheets.xlsx", sheet = 1)
#CashFlows <- as.data.frame(t(CashFlows))
original <- CashFlows
tuition <- CashFlows[c(2,18),]
names(CashFlows)
CashFlows <- subset(CashFlows, Nets == "1")
CashFlows <- rbind(CashFlows, tuition)
```

```{r}
train <- CashFlows[,2:4]
test <- CashFlows[,c(5:6)]
train <- as.data.frame(t(train))
test <- as.data.frame(t(test))
```
```{r}
seed.val <- 1234567
set.seed(seed.val)
```

## linear model

```{r}
fit.1 <- lm(V12 ~V1, data = train)
summary(fit.1)

fit.2 <- lm(V12 ~V2, data = train)
summary(fit.2)

fit.3 <- lm(V12 ~V3, data = train)
summary(fit.3)

fit.4 <- lm(V12 ~V4, data = train)
summary(fit.4)

fit.5 <- lm(V12 ~V5, data = train)
summary(fit.5)

fit.6 <- lm(V12 ~V6, data = train)
summary(fit.6)

fit.7 <- lm(V12 ~V7, data = train)
summary(fit.7)

fit.8 <- lm(V12 ~V8, data = train)
summary(fit.8)

fit.9 <- lm(V12 ~V9, data = train)
summary(fit.9)

fit.10 <- lm(V12 ~V10, data = train)
summary(fit.10)

fit.11 <- lm(V12 ~V11, data = train)
summary(fit.11)

fit.13 <- lm(V12 ~V13, data = train)
summary(fit.13)
```
note that all together, a proper linear could not be created using all the variables 

V3, V5, significant under 5% interval and V11 is significant under 1% interval. So we will use these three models of linear regresssion to try to predict tuittion and fees

V11 - sumplimental disclosures of non cash activities
V5 - net decrease in cash and noncash equivalents
V3 - cash flows from capital and other financing activities

We also found that cash from non financing activities is important to umass since that is what they are using for refunds this semester
```{r}

x <- original[c(2,3,16:20),]
x.train <- x[,2:4]
x.train <- as.data.frame(t(x.train))

x.2 <- lm(V1~V2,data=x.train)
summary(x.2)

x.3 <- lm(V1~V3,data=x.train)
summary(x.3)

x.4 <- lm(V1~V4,data=x.train)
summary(x.4)

x.5 <- lm(V1~V5,data=x.train)
summary(x.5)

x.6 <- lm(V1~V6,data=x.train)
summary(x.6)

x.7 <- lm(V1~V7,data=x.train)
summary(x.7)

```
 V7  and V2 are the only significant variables in this catergory so we will also use it
 V2 - grants and contracts
 v7 - student orginzation agency transactions
 
 
 plot tuitiin and fees on its , if you want we can plot the other variables but that would be a lot of graphs
```{r}
tuition <- as.data.frame(t(CashFlows))
 tuition.ts <- ts((as.numeric(as.vector(tuition[2:6,12]))), frequency = 1, start = c(2015, 1))
ts.plot(tuition.ts, xlab = "year", ylab ="amount in 1000 of dollars", main = "tuition over the years")
```
make our predictions
```{r}
pred.v11 <- predict(fit.11, train)
pred.v5 <- predict(fit.5, train)
predict.v3 <- predict(fit.3, train)
predict.x2 <- predict(x.2, train)
predict.x7 <- predict(x.7, train)

train.ts <- ts((as.numeric(as.vector(train$V12))), frequency = 1, start = c(2015, 1))

v11.ts <- ts(as.numeric(pred.v11), frequency = 1, start = c(2015, 1))
v5.ts <- ts(as.numeric(pred.v5), frequency = 1, start = c(2015, 1))
v3.ts <- ts(as.numeric(predict.v3), frequency = 1, start = c(2015, 1))
x2.ts <- ts(as.numeric(predict.x2), frequency = 1, start = c(2015, 1))
x7.ts <- ts(as.numeric(predict.x7), frequency = 1, start = c(2015, 1))

ts.plot(v11.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)
ts.plot(v5.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)
ts.plot(v3.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)
ts.plot(x2.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)
ts.plot(x7.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)

``` 
we have to add some labels lol but we can work on it after

v11 is by far the closest predictors, follows by v5, v3 is meh and everything else goes to shit from there

we will now plot the residual time series see how much we are off by

```{r}
rmse(pred.v5, train$V12) #12532.33
rmse(predict.v3, train$V12) #6991.26
rmse(pred.v11, train$V12) #2333.271
rmse(predict.x2, train$V12) #142085.6
rmse(predict.x7, train$V12) #42638079

```

```{r}
res.v6 = pred.v5 - train$V12
res.v11 = pred.v11 - train$V12
res.v3 = predict.v3 - train$V12
res.x2 = predict.x2 - train$V12
res.x7 = predict.x7 - train$V12

sum(res.v6) #-3.49246e-10
sum(res.v11)# 0
sum(res.v3)# 3.49246e-10
sum(res.x2)# -356311 # the last two are really bad
sum(res.x7) #124829573
```
```{r}
v6.res.ts <- ts((as.numeric(as.vector(res.v6))), frequency = 1, start = c(2015, 1))
ts.plot(v6.res.ts, xlab = "Year", ylab = "Cost in $1000", main= "Linear Model")

v11.res.ts <- ts((as.numeric(as.vector(res.v11))), frequency = 1, start = c(2015, 1))
ts.plot(v11.res.ts, xlab = "Year", ylab = "Cost in $1000", main= "Linear Model")

v3.res.ts <- ts((as.numeric(as.vector(res.v3))), frequency = 1, start = c(2015, 1))
ts.plot(v3.res.ts, xlab = "Year", ylab = "Cost in $1000", main= "Linear Model")

x2.res.ts <- ts((as.numeric(as.vector(res.x2))), frequency = 1, start = c(2015, 1))
ts.plot(x2.res.ts, xlab = "Year", ylab = "Cost in $1000", main= "Linear Model")

x7.res.ts <- ts((as.numeric(as.vector(res.x7))), frequency = 1, start = c(2015, 1))
ts.plot(x7.res.ts, xlab = "Year", ylab = "Cost in $1000", main= "Linear Model")
 
```
 all centers at 0 but last 2, the second seems to have the smallest range which means its the best, so lets use the first three as our test
 
##random forest

random forest on all of cash flows nets, var imp plot doesnt want to work so we will address that  in the plot
```{r}
train.forest <- randomForest(V12 ~.,data = train, mtry  = 4, importance = T)
#varImpPlot(train.forest)
train.pred <- predict(train.forest, train)
train.forest.pred.ts <- ts((as.numeric(as.vector(train.pred))), frequency = 1, start = c(2015, 1))
train.ts <- ts((as.numeric(as.vector(train$V12))), frequency = 1, start = c(2015, 1))

ts.plot(train.forest.pred.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)
```

```{r}
#par(mfrow= c(1,2))
train.rmse <- rmse(train$V12, train.pred)
#23,161,920.7
train.res = train.pred - train$V12
sum(train.res) # 14313.16
train.res.ts <- ts((as.numeric(as.vector(train.res))), frequency = 1, start = c(2015, 1))
ts.plot(train.res.ts, xlab = "Year", ylab = "Cost in $", main = "residuals")

#hist(train.res)
```

alot of varaiance so lets test on x

```{r}
x.forest <- randomForest(V1 ~.,data = x.train, mtry  = 5, importance = T)
#varImpPlot(train.forest)
x.pred <- predict(x.forest, x.train)
x.pred.ts <- ts((as.numeric(as.vector(x.pred))), frequency = 1, start = c(2015, 1))
#train.ts <- ts((as.numeric(as.vector(train$V12))), frequency = 1, start = c(2015, 1))

ts.plot(x.pred.ts, train.ts, gpars = list(col=rainbow(2))) 
legend("topright", legend = c("predictions", "true"), col = rainbow((2)), lty = 1)
```
```{r}
#par(mfrow= c(1,2))
x.rmse <- rmse(train$V12, x.pred)
#24,093.50
x.res = x.pred - train$V12
sum(x.res) # 8214.76 
x.res.ts <- ts((as.numeric(as.vector(x.res))), frequency = 1, start = c(2015, 1))
ts.plot(x.res.ts, xlab = "Year", ylab = "Cost in $", main = "residuals")

#hist(train.res)
```
